The core transcriptome of mammalian placentas and the divergence of expression with placental shape

### 2016-11-18 manuscript/manuscript.Rnw
switch to single-column style
reduce length of manuscript some more

%\documentclass[english,12pt]{article}
\documentclass[5p]{elsarticle_dla}
\usepackage{fontspec}
\setmainfont{FreeSerif}
\setsansfont{FreeSans}
\setmonofont{FreeMono}
\setmathrm{FreeSerif}
\setmathsf{FreeSans}
\setmathtt{FreeSerif}
%\usepackage[letterpaper,left=1in,right=1in,top=1in,bottom=1in,headheight=14pt]{geometry}
%\usepackage[letterpaper]{geometry}
%\usepackage{url}
\usepackage{graphicx}
%\usepackage{rotating}
%\usepackage{wrapfig}
\usepackage{booktabs}
%\usepackage{float}
\usepackage{multirow}
%\usepackage{setspace}
\usepackage{acro}
\usepackage{array}
\usepackage{longtable}
\usepackage{amsmath}
\usepackage[backend=biber,natbib=true,hyperref=true,citestyle=numeric-comp,style=nature,autocite=inline]{biblatex}
\addbibresource{references.bib}
\usepackage[hyperfigures,bookmarks,colorlinks,linkcolor=black,urlcolor=black,citecolor=blue,breaklinks=true]{hyperref}
% Floats at end for journals
% \usepackage[nolists,figuresfirst,nomarkers,heads]{endfloat}
% \renewcommand{\efloatseparator}{\mbox{}}
%\AtBeginFigures{\newpage \renewcommand{\efloatseparator}{\mbox{}}\section*{Figure Legends}}
%\AtBeginTables{\newpage \section*{Tables}\renewcommand{\efloatseparator}{\mbox{}}}
% \renewcommand{\includegraphics}[2][]{}
\usepackage[nomargin,inline,draft]{fixme}
\newcommand{\DLA}[1]{\textcolor{red}{\fxnote{DLA: #1}}}
\newcommand{\DEW}[1]{\textcolor{purple}{\fxnote{DEW: #1}}}
% \usepackage[x11names,svgnames]{xcolor}
% \usepackage{texshade}
% \usepackage{tikz}
\usepackage{nameref}
\usepackage{xspace}
% this package will stretch the font slightly if necessary.
% \usepackage{microtype}
\usepackage{zref-xr,zref-user}
\zxrsetup{toltxlabel}
\zexternaldocument*{supplemental_material}
\usepackage[capitalize]{cleveref}
\def\UrlBreaks{\do\.\do\@\do\\\do\/\do\!\do\_\do\|\do\;\do\>\do\]%
\do\)\do\,\do\?\do\&\do\'\do+\do\=\do\#%
\do\/%
}%
\input{common_latex}
\tolerance=5000
\emergencystretch=0.1pt

@@

Previous studies of some species (i.e., \textit{Homo sapiens},
\textit{Mus musculus}, \textit{Bos taurus}) have provided some
preliminary answers to these questions. They revealed a stunning
divergence in the genetics of the placenta, especially in later stages
of pregnancy
\cite{Carter2012:Evolutionofplacentalfunction,Knox.ea2008:Genomicevolutionofplacenta,Su.ea2002:Large-scaleanalysisofhuman}.
For example, each lineage has evolved expansive placenta-specific gene
families with expression patterns that can vary in space and time
\cite{Green.ea2000:Pregnancy-associatedbovineandovine}.
These include independent expansions of the prolactins in \mouse,
\textit{Spalax carmeli}, \textit{Spalax galili}, and \cow, \acp{PSG}
in \human and \mouse, \acp{PAG} in \cow, genes related to the
thymus-specific serine protease \textit{PRSS16} in \textit{Loxodona
  africana} (in this work), and the expansion of the growth hormone
and chorionic gonadotropin beta genes in anthropoid primates
\cite{Haig2008:Placentalgrowthhormone-relatedproteins,Rawn.ea2008:evolutionregulationandfunction}.
Expression patterns of specific genes also can vary widely between
species. For example, the expression of orthologous genes in the
placenta have been found to vary substantially between \human and
\mouse \cite{Su.ea2002:Large-scaleanalysisofhuman}, and the
recruitment of \textit{SIGLEC6} to placental expression in \human
differs from both \textit{Pan troglodytes} and \mouse
\cite{Linden.ea2007:Human-specificexpressionofSiglec-6}.

In order to address our two questions in a more systematic way, we
sequenced and quantified the total RNA of mammalian placentas at term,
including that of the \human, \textit{Pan paniscus} (the sister
species to \textit{P. troglodytes}), \textit{Ateles fusciceps},
\mouse, \galili, \carmeli, \textit{Canis familiaris}, \cow, \elephant,
\textit{Dasypus novemcinctus}, and \textit{Monodelphis domesticus}.
These species represent four major branches of eutherian mammals
(Euarchontoglires, Laurasiatheria, Afrotheria, and Xenartha) as well
as marsupials. These placentas also represent a wide array of
placental morphologies as well as traits such as number and size of
offspring and length of gestation (\cref{tab:gestational_periods}). We
use this data, along with existing RNA-Seq data from \mouse
\cite{Wang.ea2011:surveyfornovelimprinted}, \human
\cite{Hughes.ea2015:Evaluatingintra-andinter-individual},
\textit{Equus caballus}
\cite{Coleman.ea2010:Structuralannotationofequine}, \textit{Ovis
  aries} \cite{Archibald.ea2010:sheepgenomereferencesequence}, and
\textit{Sus scrofa} \cite{Ahanda.ea2012:Predictionofaltered3-} to: 1)
identify the non-housekeeping genes which are expressed in the
placentas of all species analyzed and therefore represent the
\emph{core functionality of the placenta}; 2) identify the genes whose
expression level changes significantly in each of the morphology
types.

@@

\begin{sloppypar}
The quality of reads was evaluated using FASTQC v0.11.2
(\url{http://www.bioinformatics.babraham.ac.uk/projects/fastqc/}).
Transcriptomes without a suitable reference genome (\smonkey, \galili,
\carmeli; italic reference in Supplemental
\cref{tab:ref_genome_aligned_species}) were assembled using Trinity
v2014-07-17 \cite{Haas.ea2013:Denovotranscriptsequence} and annotated
using Diamond \cite{Buchfink.ea2015:Fastandsensitiveprotein} against
either the \human or \mouse transcriptome (Supplemental
\cref{tab:ref_genome_aligned_species}). For annotation using Diamond,
the match with the highest bitscore was chosen as the annotation;
transcripts where the highest scoring match had a bitscore below 40
were left unannotated.
% 
All transcriptomes were then aligned to the appropriate reference (or
assembled) genome using STAR v2.4.2a~%
\cite{Dobin.ea2013:STARultrafastuniversalRNA-seq} (Supplemental
\cref{fig:transcriptome_statistics,fig:transcriptome_reads_vs_read_length})
and expression values in \ac{FPKM}
% \cite{Trapnell.ea2010:Transcriptassemblyandquantification}
were
quantified using Cufflinks v2.2.1
\cite{Trapnell.ea2009:TopHatdiscoveringsplicejunctions} (Supplemental
\cref{tab:all_ateles_fusciceps_placenta,tab:all_homo_sapiens_placenta,tab:all_pan_paniscus_placenta,tab:all_mus_musculus_placenta,tab:all_spalax_galili_placenta,tab:all_spalax_carmeli_placenta,tab:all_bos_taurus_placenta,tab:all_ovis_aries_placenta,tab:all_sus_scrofa_placenta,tab:all_equus_caballus_placenta,tab:all_canis_familiaris_placenta,tab:all_loxodonta_africana_placenta,tab:all_dasypus_novemcinctus_placenta,tab:all_monodelphis_domestica_placenta}).
%
The genome of the closely-related \textit{P. troglodytes} was used as
a reference for \bonobo. The separation between the two recognized Pan
species is generally perceived to be $\approx 1$ million years
\cite{Prüfer.ea2012:bonobogenomecomparedwith}. %
%
\end{sloppypar}

@@

All 1:1 orthologs present among \opossum, \textit{Pan troglodytes},
\cow, \dog, \elephant, \mouse, and \human were determined using
Ensembl v.80 gene trees via Biomart using the biomaRt R package
\cite{Durinck.ea2009:Mappingidentifiersforintegration}. %
A total of \Sexpr{combined.fpkm[species!="homo sapiens" & mean_fpkm >=
  1,length(unique(human_name))]} 1:1 orthologous genes were expressed
in at least one species. Of these, \Sexpr{combined.fpkm[species!="homo
  sapiens" & mean_fpkm >= 100,length(unique(human_name))]} 1:1
orthologous genes with \ac{FPKM} $\ge 100$ were retained for future
analyses of gene expression evolution described below.

@@

Gene sets and pathways in were obtained from MSigdb v5.1 hallmark gene
set \cite{Liberzon.ea2011:MolecularsignaturesdatabaseMSigDB} and Gene
Ontology which represent a superset of available pathway databases,
including Reactome, Kegg, Netpath, and GO. We calculated the
over-representation of gene lists in these gene sets using Fisher's
exact test and corrected for multiple testing by estimating the
\ac{FDR} using the Benjamini and Hochberg procedure
\cite{Benjamini1995:bh_procedure}.

@@

All of the code used to produce every analysis presented in this paper
is available at
\url{https://github.com/uiuc-cgm/placenta-viviparous.git}. All of the
sequences used are deposited in SRA, and new sequences for this work
can be obtained under the series
\href{http://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE79121}{GSE79121}
(Supplemental \cref{tab:srr_numbers}). The mean FPKM for gene in each
species with ortholog annotations are located in the supplemental data
file all\_species\_mean\_fpkm.txt.xz and the per-sample FPKM values are in
all\_species\_per\_sample\_fpkm.txt.xz.

@@

We analyzed the sequenced transcriptome of
\Sexpr{star.logs[,length(unique(file))]} placentas from
\Sexpr{star.logs[,length(unique(species))]} species spanning the
phylogenetic tree of mammals. This wealth of data revealed both
lineage-specific expansions and differential expression of gene
families as well as a set of core, non-housekeeping genes which are
critical for the proper organization and functioning of the placenta.

There is diversity in the placental expression of hormones, including
estrogen, progesterone, prolactin, relaxin, the growth hormone family, and
natriuretic peptides among the species and lineages analyzed.
We also identified genes differentially expressed on the \human
lineage which are implicated in pre-eclampsia, including
\textit{ADAM12}, \textit{CRH}, and \textit{KISS1}.

The core placenta transcriptome was also identified, which contains
genes which are involved in maternal-fetal tolerance, such as the
annexins, and genes responsible for cell-cell and cell-matrix
interactions, such as \textit{LGALS1}. These genes were almost
certainly present in the placenta of the common mammalian ancestor. We
hypothesize that modifications to their expression or function will
compromise the development and organization of the placenta.

>>>>

%\documentclass[english,12pt]{article}
%\documentclass[5p]{elsarticle_dla}
\documentclass{elsarticle_dla}
\usepackage{fontspec}
\setmainfont{FreeSerif}
\setsansfont{FreeSans}
\setmonofont{FreeMono}
\setmathrm{FreeSerif}
\setmathsf{FreeSans}
\setmathtt{FreeSerif}
%\usepackage[letterpaper,left=1in,right=1in,top=1in,bottom=1in,headheight=14pt]{geometry}
%\usepackage[letterpaper]{geometry}
%\usepackage{url}
\usepackage{graphicx}
%\usepackage{rotating}
%\usepackage{wrapfig}
\usepackage{booktabs}
%\usepackage{float}
\usepackage{multirow}
%\usepackage{setspace}
\usepackage{acro}
\usepackage{array}
\usepackage{longtable}
\usepackage{amsmath}
\usepackage[backend=biber,natbib=true,hyperref=true,citestyle=numeric-comp,style=nature,autocite=inline]{biblatex}
\addbibresource{references.bib}
\usepackage[hyperfigures,bookmarks,colorlinks,linkcolor=black,urlcolor=black,citecolor=blue,breaklinks=true]{hyperref}
% Floats at end for journals
% \usepackage[nolists,figuresfirst,nomarkers,heads]{endfloat}
% \renewcommand{\efloatseparator}{\mbox{}}
%\AtBeginFigures{\newpage \renewcommand{\efloatseparator}{\mbox{}}\section*{Figure Legends}}
%\AtBeginTables{\newpage \section*{Tables}\renewcommand{\efloatseparator}{\mbox{}}}
% \renewcommand{\includegraphics}[2][]{}
\usepackage[nomargin,inline,draft]{fixme}
\newcommand{\DLA}[1]{\textcolor{red}{\fxnote{DLA: #1}}}
\newcommand{\DEW}[1]{\textcolor{purple}{\fxnote{DEW: #1}}}
% \usepackage[x11names,svgnames]{xcolor}
% \usepackage{texshade}
% \usepackage{tikz}
\usepackage{nameref}
\usepackage{xspace}
% this package will stretch the font slightly if necessary.
% \usepackage{microtype}
\usepackage{zref-xr,zref-user}
\zxrsetup{toltxlabel}
\zexternaldocument*{supplemental_material}
\usepackage[capitalize]{cleveref}
\def\UrlBreaks{\do\.\do\@\do\\\do\/\do\!\do\_\do\|\do\;\do\>\do\]%
\do\)\do\,\do\?\do\&\do\'\do+\do\=\do\#%
\do\/%
}%
\input{common_latex}
\tolerance=5000
\emergencystretch=0.1pt

@@

Previous studies of some species (i.e., \textit{Homo sapiens},
\textit{Mus musculus}, \textit{Bos taurus}) have provided some
preliminary answers to these questions. They revealed a stunning
divergence in the genetics of the placenta, especially in later stages
of pregnancy
\cite{Carter2012:Evolutionofplacentalfunction,Knox.ea2008:Genomicevolutionofplacenta,Su.ea2002:Large-scaleanalysisofhuman}.
For example, each lineage has evolved expansive placenta-specific gene
families with expression patterns that can vary in space and time
\cite{Green.ea2000:Pregnancy-associatedbovineandovine}.
These include independent expansions of the prolactins in \mouse,
\textit{Spalax carmeli}, \textit{Spalax galili}, and \cow, \acp{PSG}
in \human and \mouse, \acp{PAG} in \cow, genes related to the
thymus-specific serine protease \textit{PRSS16} in \textit{Loxodona
  africana} (in this work), and the expansion of the growth hormone
and chorionic gonadotropin beta genes in anthropoid primates
\cite{Haig2008:Placentalgrowthhormone-relatedproteins,Rawn.ea2008:evolutionregulationandfunction}.
Expression patterns of specific genes also can vary widely between
species. For example, the expression of orthologous genes in the
placenta have been found to vary substantially between \human and
\mouse \cite{Su.ea2002:Large-scaleanalysisofhuman}.

To address our questions, we 
sequenced and quantified the total RNA of mammalian placentas at term,
including \human, \textit{Pan paniscus} (the sister
species to \textit{P. troglodytes}), \textit{Ateles fusciceps},
\mouse, \galili, \carmeli, \textit{Canis familiaris}, \cow, \elephant,
\textit{Dasypus novemcinctus}, and \textit{Monodelphis domesticus}.
These species represent four major branches of eutherian mammals
(Euarchontoglires, Laurasiatheria, Afrotheria, and Xenartha) as well
as marsupials. These placentas also represent a wide array of
placental morphologies as well as traits such as number and size of
offspring and length of gestation (\cref{tab:gestational_periods}). We
use this data, along with existing RNA-Seq data from \mouse
\cite{Wang.ea2011:surveyfornovelimprinted}, \human
\cite{Hughes.ea2015:Evaluatingintra-andinter-individual},
\textit{Equus caballus}
\cite{Coleman.ea2010:Structuralannotationofequine}, \textit{Ovis
  aries} \cite{Archibald.ea2010:sheepgenomereferencesequence}, and
\textit{Sus scrofa} \cite{Ahanda.ea2012:Predictionofaltered3-} to: 1)
identify the non-housekeeping genes which are expressed in the
placentas of all species analyzed and therefore represent the
\emph{core functionality of the placenta}; 2) identify genes whose
expression level changes significantly in each of the morphology
types.

@@

\begin{sloppypar}
Reads quality was evaluated using FASTQC v0.11.2
(\url{http://www.bioinformatics.babraham.ac.uk/projects/fastqc/}).
Transcriptomes without a suitable reference genome (\smonkey, \galili,
\carmeli; italic reference in Supplemental
\cref{tab:ref_genome_aligned_species}) were assembled using Trinity
v2014-07-17 \cite{Haas.ea2013:Denovotranscriptsequence} and annotated
using Diamond \cite{Buchfink.ea2015:Fastandsensitiveprotein} against
either the \human or \mouse transcriptome (Supplemental
\cref{tab:ref_genome_aligned_species}), and the
the match with the highest bitscore was chosen as the annotation.
Transcripts where the highest scoring match had a bitscore below 40
were left un-annotated.
% 
All transcriptomes were aligned to the reference (or
assembled) genome using STAR v2.4.2a~%
\cite{Dobin.ea2013:STARultrafastuniversalRNA-seq} (Supplemental
\cref{fig:transcriptome_statistics,fig:transcriptome_reads_vs_read_length})
and expression values in \ac{FPKM}
% \cite{Trapnell.ea2010:Transcriptassemblyandquantification}
were
quantified using Cufflinks v2.2.1
\cite{Trapnell.ea2009:TopHatdiscoveringsplicejunctions} (Supplemental
\cref{tab:all_ateles_fusciceps_placenta,tab:all_homo_sapiens_placenta,tab:all_pan_paniscus_placenta,tab:all_mus_musculus_placenta,tab:all_spalax_galili_placenta,tab:all_spalax_carmeli_placenta,tab:all_bos_taurus_placenta,tab:all_ovis_aries_placenta,tab:all_sus_scrofa_placenta,tab:all_equus_caballus_placenta,tab:all_canis_familiaris_placenta,tab:all_loxodonta_africana_placenta,tab:all_dasypus_novemcinctus_placenta,tab:all_monodelphis_domestica_placenta}).
%
The genome of the closely-related \textit{P. troglodytes} (separation
$\approx 1$ million years
\cite{Prüfer.ea2012:bonobogenomecomparedwith}) was used as a reference
for \bonobo %
.
\end{sloppypar}

@@

1:1 orthologs in were determined using
Ensembl v.80 gene trees via Biomart using the biomaRt R package
\cite{Durinck.ea2009:Mappingidentifiersforintegration}. %
A total of \Sexpr{combined.fpkm[species!="homo sapiens" & mean_fpkm >=
  1,length(unique(human_name))]} 1:1 orthologous genes were expressed
in at least one species.

@@

Gene sets and pathways in were obtained from MSigdb v5.1
\cite{Liberzon.ea2011:MolecularsignaturesdatabaseMSigDB} and Gene
Ontology to form a superset of available pathway databases. We
calculated the over-representation of gene lists in these gene sets
using Fisher's exact test and corrected for multiple testing by
estimating the \ac{FDR} using the Benjamini and Hochberg procedure
\cite{Benjamini1995:bh_procedure}.

@@

All of the code used to produce every analysis presented in this paper
is available at
\url{https://github.com/uiuc-cgm/placenta-viviparous.git}. All of the
sequences used are deposited in SRA, and new sequences for this work
can be obtained under the series
\href{http://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE79121}{GSE79121}
(Supplemental \cref{tab:srr_numbers}). The mean \ac{FPKM} for gene in each
species with ortholog annotations are located in the supplemental data
file all\_species\_mean\_fpkm.txt.xz and the per-sample \ac{FPKM} values are in
all\_species\_per\_sample\_fpkm.txt.xz.

@@

We analyzed the sequenced transcriptome of
\Sexpr{star.logs[,length(unique(file))]} placentas from
\Sexpr{star.logs[,length(unique(species))]} species spanning the
phylogenetic tree of mammals. This revealed both
lineage-specific differential expression of gene
families as well as a set of core, non-housekeeping genes which are
critical for the proper organization and functioning of the placenta.

The core placenta transcriptome contains genes which are involved in
maternal-fetal tolerance, such as the annexins, and genes responsible
for cell-cell and cell-matrix interactions, such as \textit{LGALS1}.
These genes were almost certainly present in the placenta of the
common mammalian ancestor. We hypothesize that modifications to their
expression or function will compromise the development and
organization of the placenta.

We also identified genes differentially expressed on the \human
lineage which are implicated in pre-eclampsia, including
\textit{ADAM12}, \textit{CRH}, and \textit{KISS1}.
